\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-10-11 papers/methods/paper.tex
More of ricks feedback

\newcommand{\kmsmpc}{{\rm km}\,{\rm s}^{-1}\,{\rm Mpc}^{-1}}
\newcommand{\cov}{\mathcal{C}}
\newcommand{\Z}{\mathcal{Z}}

@@

Two decades have passed since the discovery of the accelerating universe \citep{Riess1998, Perlmutter1999}. Since that time, the number of observed type Ia supernovae (SN~Ia) has increased by more than an order of magnitude, with contributions from modern surveys at both low redshift \citep{Bailey2008, Freedman2009, Hicken2009,  Contreras2010, Conley2011}, and higher redshift \citep{Astier2006, Wood-Vasey2007, Frieman2008, Balland2009, Amanullah2010, chambers2016panstarrs, sako2018sdss}. Cosmological analyses of these supernova samples \citep{Kowalski2008, Kessler2009, Conley2011, Suzuki2012, Betoule2014, Rest2014, Scolnic2017} have been combined with complementary probes of large scale structure and the CMB. For a recent review, see \citet{Huterer2018}. {\green While these efforts have reduced the uncertainty on the equation-of-state of dark energy by more than a factor of two, it is still consistent with a cosmological constant and the nature of dark energy remains an unsolved mystery.}

In attempts to tease out the nature of dark energy, active and planned surveys are {\green continually growing in size and scale}. The Dark Energy Survey \citep[DES,][]{Bernstein2012, Abbott2016} has {\green discovered} thousands of type Ia supernovae, attaining both spectroscopically and photometrically identified samples. The Large Synoptic Survey Telescope \citep[LSST,][]{Ivezic2008, LSSTScienceCollaboration2009} will {\green discover} scores of thousands of photometrically classified supernovae. {\green Such increased statistical power demands greater fidelity and flexibility in modelling supernovae for cosmological purposes, as we will require reduced systematic uncertainties to fully utilise these increased statistics} \citep{Betoule2014, Scolnic2017}.

As such, {\green considerable resources are} being put into developing more sophisticated supernova cosmology analyses. The role of simulations mimicking survey observations has become increasingly important in determining biases in cosmological constraints and validating specific supernova models. First used in SNLS \citep{Astier2006} and ESSENCE analyses \citep{Wood-Vasey2007}, and then refined and improved in \citet{Kessler2009}, simulations are now a fundamental component of modern supernova cosmology.  \citet{Betoule2014} quantise and then correct observational bias using simulations, and more recently \citet{Scolnic2016} and \citet{Kessler2017} explore simulations to quantify observational bias in SN Ia distances as a function of {\green multiple} factors to better remove biases. Approximate Bayesian computation methods also make use of simulations, trading traditional likelihoods and analytic approximations for more robust models with the cost of increased computational time \citep{Weyant2013, Jennings2016}. Bayesian Hierarchical models abound \citep{Mandel2009, March2011, March2014, Rubin2015, Shariff2016, Roberts2017}, {\green and either use simulation-determined distance-corrections} or attempt to find analytic approximations for complicated effects such as Malmquist bias.

{\green In this paper, we lay out a new hierarchical model that builds off the past work of \citet{Rubin2015}. We include additional sources of systematic uncertainty, including an analytic formulation of selection efficiency which incorporates parametric uncertainty. We also implement a different model of intrinsic dispersion to both incorporate redshift-dependent scatter and to increase numerical efficiency, allowing our model to perform rapid fits to supernovae datasets.} Section \ref{sec:review} is dedicated to a quick review of {\green supernova cosmology analysis methods}, and Section \ref{sec:challenges} outlines some of the common challenges faced by analysis methods. In Section \ref{sec:method} we outline our methodology. Model verification on simulated datasets is given in Section \ref{sec:verification}, along with details on potential areas of improvement. We summarise our methodology in Section \ref{sec:conclusion}.

@@

Whilst supernova observations take the form of photometric time-series brightness measurements in many bands {\green and a redshift measurement of the supernova (or its assumed host)}, most analyses do not work from these measurements directly. Instead, most techniques fit an observed redshift and these photometric observations to a supernova model, with the most widely used being that of the empirical SALT2 model \citep{Guy2007, Guy2010}. This model is trained separately before fitting the supernova light curves for cosmology \citep{Guy2010, Betoule2014}. The resulting output from the model is, for each supernova, an amplitude $x_0$ (which can be converted into apparent magnitude, $m_B = -2.5\log(x_0)$), a stretch term $x_1$, and color term $c$, along with a covariance matrix describing the uncertainty on these summary statistics ($\mathcal{C}$). As all supernovae are not identical, an ensemble of supernovae form a redshift-dependent, observed population of $\hat{m}_B$, $\hat{x}_1$ and $\hat{c}$, where the hat denotes an observed variable.

This {\green ensemble of $\hat{m}_B$, $\hat{x}_1$ and $\hat{c}$} represents an observed population, which -- due to the presence of various selection effects -- may not represent the true, underlying supernova population. Accurately characterising this underlying population, its evolution over redshift, and effects from {\green host-galaxy} environment, is one of the challenges of supernova cosmology. Given some modelled underlying supernova population that lives in the redshift-dependent space $M_B$ {\green(absolute magnitude of the supernova, traditionally in the Bessell $B$ band)}, $x_1$, and $c$, the introduction of cosmology into the model is simple -- it translates the underlying population from absolute magnitude space into the observed population in apparent magnitude space. Specifically, for any given supernova our map between absolute magnitude and apparent magnitude may take the traditional form:

\begin{align}
\mu(z) = m_B + \alpha x_1 - \beta c - M_B + \text{other corrections}, \label{eq:standard}
\end{align}

where $M_B$ is the mean absolute magnitude for all SN Ia,  $\alpha$ is the stretch correction \citep{Phillips1993, Phillips1999}, and $\beta$ is the color correction \citep{Tripp1998} that respectively encapsulate the empirical relation that broader (longer-lasting) and bluer supernovae are brighter. The cosmological term, $\mu(z)$ represents the distance modulus, and can be precisely calculated given cosmological parameters and a redshift, or can be `observed' by {\green utilising equation \eqref{eq:standard} and supernova observations}. The `$\text{other corrections}$' term often includes corrections for host galaxy environment, as this has statistically significant correlations with supernova properties \citep{Kelly2010, Lampeitl2010, Sullivan2010, DAndrea2011, Gupta2011, Johansson2013, Rigault2013, Uddin2017}. In traditional analyses, Malmquist bias corrections are also included, which can take the form of a redshift-dependent function \citep{Betoule2014} or a 5D function of $c$, $x_1$, $\alpha$, $\beta$ and $z$ \citep{Scolnic2017}.

@@

Traditional $\chi^2$ analyses such as that found in \citet{Riess1998, Perlmutter1999, Wood-Vasey2007, Kowalski2008, Kessler2009, Conley2011, Betoule2014}, minimise the difference in distance modulus between the observed distance modulus attained from equation \ref{eq:standard}, and the cosmologically predicted values. We denote the observed distance modulus $\mu_{\rm obs}$ and the cosmologically calculated one $\mu_C$. {\green The $\chi^2$ function minimised is}

\begin{align}
\chi^2 = (\mu_{\rm obs} - \mu_C)^\dagger C^{-1} (\mu_{\rm obs} - \mu_C),
\end{align}

where $C^{-1}$ is an uncertainty matrix that combines statistical and systematic uncertainties (see \citet{Brout18SYS} for a review of these uncertainties). The cosmological $\mu_C$ is defined as
{\green 
\begin{align}
\mu_C &= 5 \log\left[ \frac{d_L}{10{\rm pc}} \right], \label{eq:mu0}\\
d_L& = (1+z)\frac{c}{H_0} \int_0^z \frac{dz'}{\sqrt{ \Omega_m (1+z')^3 + \Omega_k (1 + z')^2 + \Omega_\Lambda (1+z')^{3(1+w)}}},
\end{align}
}
where $d_L$ is the luminosity distance for redshift $z$ given a specific cosmology, $H_0$ is the current value of Hubble's constant in $\kmsmpc$ and $\Omega_m$, $\Omega_k$, and $\Omega_\Lambda$ represent the energy density terms for matter, curvature and dark energy respectively.

The benefit this analysis methodology provides is speed -- for samples of hundreds of supernovae or less, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with several limitations. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of Gaussian uncertainty. Secondly, the method of creating a global covariance matrix relies on computing partial derivatives and thus any uncertainty estimated from this method loses all information about correlation between sources of uncertainty. For example, the underlying supernova color population's mean and skewness are highly correlated, however this correlation is lost when determining population uncertainty using numerical derivatives of population permutations. {\green Whilst known correlations like this can be captured and incorporated into a covariance matrix if known, it requires human awareness of the correlations and thus methods which can automatically capture correlated uncertainties are preferable.} Thirdly, the computational efficiency is dependent on both creating the global covariance matrix, and then inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae. A recent solution to this computational cost problem is to bin the supernovae {\green in redshift bins}, which then produces a matrix of arbitrary size {\green and can allow fitting without matrix inversion on every step}. Whilst binning data results in some loss of information, recent works tested against simulations show that this loss of information does not create significant cosmological biases \citep{Scolnic2016, Kessler2017}.

Selection efficiency, such as the well known Malmquist bias \citep{MalmquistK.G.1922} is accounted for by correcting data. Specifically, Malmquist bias is the result of losing the fainter tail of the supernova population at high redshift when supernovae increasingly fall below the detection threshold. An example of Malmquist bias is illustrated in Figure \ref{fig:malmquist}. Simulations following survey observational strategies and geometry are used to calculate the expected bias in distance modulus, which is then {\green subtracted from} the observational data. When using traditional fitting methods such as that found in \citet{Betoule2014}, these effects are not built into the likelihood and instead are formed by correcting data, which means their uncertainty is not captured fully in the $\chi^2$ distribution, and any subtle correlations between cosmological or population parameters and the bias is lost. {\green Recent developments such as the BBC method \citep{Kessler2017} incorporate corrections dependent on $\alpha$ and $\beta$, improving their capture of uncertainty on bias corrections.}

@@

Sitting between the traditional models simplicity and the complexity of forward modelling lies Bayesian hierarchical models (BHM). Hierarchical models utilise multiple layers of connected parameters, with the layers linked via well defined and {\green physically} motivated conditional probabilities. For example, an observation of a parameter from a population will be conditioned on the true value of the parameter, which itself will be conditioned on the population distribution of that parameter. We can thus easily incorporate different distributions, populations, and parameter inter-dependence which cannot be found in traditional analyses where uncertainty must be encapsulated in a covariance matrix.

With the introduction of multiple layers in our model, we can add more flexibility than a traditional analysis whilst still maintaining most of the computational benefits that come from having a tractable likelihood. \citet{Mandel2009, Mandel2011a, Mandel2017} construct a hierarchical model that they apply to supernova light curve fitting. \citet{March2011} derive a hierarchical model and simplify it by analytically marginalising over nuisance parameters to provide increased flexibility with reduced uncertainty over the traditional method, but do not incorporate bias correction. \citet{March2014, Karpenka2015} improve upon this by incorporating redshift-dependent magnitude corrections from \citet{Perrett2010} to remove bias, and validate on 100 realisations of SNLS like simulations. The recent BAHAMAS model \citep{Shariff2016} builds on this and reanalyses the JLA dataset \citep[using redshift dependent bias corrections from][]{Betoule2014}, whilst including extra freedom in the correction factors $\alpha$ and $\beta$, finding evidence for redshift dependence on $\beta$. \citet{Ma2016} performed a reanalysis of the JLA dataset within a Bayesian formulation, finding significant differences in $\alpha$ and $\beta$ values from the original analysis from \citet{Betoule2014}. Notably, these methods rely on data that is bias corrected or ignore biases, however the UNITY framework given by \citet{Rubin2015} incorporates selection efficiency analytically in the model, and is applied to the Union 2.1 dataset \citep{Suzuki2012}. The simplification made by the UNITY analysis is that the bias is {\green perfectly described} by an analytic function. Their model was validated to be free of significant biases using fits to {\green thirty realisations of supernova datasets that are constructed to mimic the UNITY framework}. The well known BEAMS (Bayesian estimation applied to multiple species) methodology from \citet{Kunz2007} has been extended and applied in several works \citep{Hlozek2012}, most lately to include redshift uncertainty for photometric redshift application as zBEAMS \citep{Roberts2017} and to include simulated bias corrections in \citet{Kessler2017}. For the latter case, by inferring biases using Bayesian models, sophisticated corrections can be calculated and then applied to more traditional $\chi^2$ models.

Whilst there are a large number of hierarchical models available, none of them have undergone high-statistic tests using realistic simulations to verify each models' respective bias. {\green Additionally, testing has generally been performed on supernovae simulations with either $\Lambda$CDM cosmology or Flat $\Lambda$CDM cosmology. However, quantifying the biases on $w$CDM cosmology simulations with realistic simulations is becoming critically important as precision supernovae cosmology comes into its own, and focus shifts from determination of $\Omega_m$ to $w$.}

The flexibility afforded by hierarchical models allows for investigations into different treatments of underlying supernova magnitude, color and stretch populations, host-galaxy corrections and redshift evolution, each of which will be discussed further in the outline of our model below. Our model is designed to increase the numerical efficiency of prior works whilst incorporating the flexibility of hierarchical models. {\green We reduce our dependence on an assumed scatter model in simulations by not utilising bias-corrections} to provide a valuable cross-check on analysis methodologies which are highly simulation-dependent.

@@

{\green Forefront among these challenges is our ignorance of the underlying type Ia population -- its intrinsic dispersion}. Ideally, analysis of the underlying population would make use of an ensemble of time-series spectroscopy to characterise the diversity of type Ia supernovae, however this data is difficult to obtain, and recent efforts to quantify the population draw inference from photometric measurements. We{\green test against }two dispersion models in this work. The {\green first is based on the} \citet[][hereafter denoted {\gten}]{Guy2010} scatter model, which models intrinsic scatter with a 70\% contribution from coherent variation and 30\% from chromatic variation. The second, {\green based on the} {\celeven} model, is sourced from \citet{Chotard2011} and has variation with 25\% contribution from coherent scatter and 75\% from chromatic variation. {\green These two broadband scatter models are converted to spectral energy distribution models for use in simulations in \citet{Kessler13}.} As the SALT2 model does not include full treatment of intrinsic dispersion, each scatter model results in different biases in $m_B$, $x_1$, and $c$ when fitting the SALT2 model to light curve observations, and results in increased uncertainty on the summary statistics that is not encapsulated in the reported covariance $\mathcal{C}$. These two scatter models are currently {\green assumed} to span the possible range of scatter in the underlying supernova population. We have insufficient information to prefer one model over the other, and thus we have to account for both possible scatter models.

The underlying supernova population is further complicated by the presence of outliers. Non-type Ia supernovae often trigger transient follow-up in surveys and can easily be mistaken for type Ia supernovae {\green and represent outliers from the standardised SN~Ia sample.} This contamination is not just a result of non-SN~Ia being observed, but can also arise from host galaxy misidentification causing incorrect redshifts. Depending on the cuts {\green optimising the} ratio between purity and efficiency, this can result in between 3\% to 9\% misidentification \citep{Gupta2016} and results in a broad population of outliers. For spectroscopic surveys, where both supernova type and redshift can be confirmed through the supernova spectra, this outlier population is negligible. However, for photometric surveys, which do not have the spectroscopic confirmation, it is one of the largest challenges; how to model, fit, and correct for contaminants.

Finally, one of the other persistent challenges facing supernova cosmology analyses are the high number of systematics. Because of the rarity of SN~Ia, datasets are commonly formed from the discoveries of multiple surveys. However, each different survey introduces additional sources of systematic error, from sources within each survey such as band calibration, to systematics introduced by calibration across surveys. Peculiar velocities, different host environments, and dust extinction represent additional sources of systematic uncertainty which must all be modelled and accounted for.

@@

We construct our hierarchical Bayesian model with several goals in mind: creation of a redshift-dependent underlying supernova population, {\green treatment of an increased number of systematics}, and analytic correction of selection effects, including systematic uncertainty on those corrections. We also desire this to be more computationally efficient than prior works, such that cosmological results from thousands of supernovae are obtainable in the order of hours, rather than days. As this is closest to the UNITY method from \citet[][hereafter denoted \rubin]{Rubin2015}, we follow a similar model scaffold, and construct the model in the programming language Stan \citep{Carpenter2017, StanDevelopmentTeam2017}. The primary challenge of fitting hierarchical models is their large number of fit parameters, and Stan, which uses automatic differentiation and the no-U-turn Sampler (NUTS, a variant of Hamiltonian Monte Carlo), allows us to efficiently sample high dimensional parameter space.

@@

\subsubsection{Observables}
Like most of the BHM methods introduced previously, we work from the summary statistics, where each observed supernova has a flux measurement $\hat{m}_B$ (which is analogous to apparent magnitude), stretch $\hat{x}_1$ and color $\hat{c}$, with uncertainty on those values encoded in the covariance matrix $\cov$. Additionally, each supernova has an observed redshift $\hat{z}$ and a host galaxy {\green stellar }mass associated with it, $\hat{m}$, where the mass measurement takes the form of a probability of being above $10^{10}$ solar masses. We will also have a probability of each supernova being a type Ia, $\hat{p}$. Our set of observables {\green input into the BHM model} is therefore given as $\lbrace \hat{m}_B, \hat{x}_1, \hat{c}, \hat{z}, \hat{p}, \hat{m}, \cov \rbrace$, as shown in the probabilistic graphical model (PGM) in Figure \ref{fig:pgm}.

>>>>

\newcommand{\kmsmpc}{{\rm km}\,{\rm s}^{-1}\,{\rm Mpc}^{-1}}
\newcommand{\cov}{\mathcal{C}_\eta}
\newcommand{\Z}{\mathcal{Z}}

@@

Two decades have passed since the discovery of the accelerating universe \citep{Riess1998, Perlmutter1999}. Since that time, the number of observed type Ia supernovae (SN~Ia) has increased by more than an order of magnitude, with contributions from modern surveys at both low redshift \citep{Bailey2008, Freedman2009, Hicken2009,  Contreras2010, Conley2011}, and higher redshift \citep{Astier2006, Wood-Vasey2007, Frieman2008, Balland2009, Amanullah2010, chambers2016panstarrs, sako2018sdss}. Cosmological analyses of these supernova samples \citep{Kowalski2008, Kessler2009, Conley2011, Suzuki2012, Betoule2014, Rest2014, Scolnic2017} have been combined with complementary probes of large scale structure and the CMB. For a recent review, see \citet{Huterer2018}. {\green While these efforts have reduced the uncertainty on the equation-of-state of dark energy ($w$) by more than a factor of two, it is still consistent with a cosmological constant and the nature of dark energy remains an unsolved mystery.}

In attempts to tease out the nature of dark energy, active and planned surveys are {\green continually growing in size and scale}. The Dark Energy Survey \citep[DES,][]{Bernstein2012, Abbott2016} has {\green discovered} thousands of type Ia supernovae, attaining both spectroscopically and photometrically identified samples. The Large Synoptic Survey Telescope \citep[LSST,][]{Ivezic2008, LSSTScienceCollaboration2009} will {\green discover tens} of thousands of photometrically classified supernovae. {\green Such increased statistical power demands greater fidelity and flexibility in modelling supernovae for cosmological purposes, as we will require reduced systematic uncertainties to fully utilise these increased statistics} \citep{Betoule2014, Scolnic2017}.

As such, {\green considerable resources are aimed at} developing more sophisticated supernova cosmology analyses. The role of simulations mimicking survey observations has become increasingly important in determining biases in cosmological constraints and validating specific supernova models. First used in SNLS \citep{Astier2006} and ESSENCE analyses \citep{Wood-Vasey2007}, and then refined and improved for {\green SDSS \citep{Kessler2009},} simulations are a fundamental component of modern supernova cosmology.  \citet{Betoule2014} quantise and correct observational bias using simulations, and more recently \citet{Scolnic2016} and \citet{Kessler2017} explore simulations to quantify observational bias in SN~Ia distances as a function of {\green multiple factors to improve bias correction}. Approximate Bayesian computation methods also make use of simulations, trading traditional likelihoods and analytic approximations for more robust models with the cost of increased computational time \citep{Weyant2013, Jennings2016}. Bayesian Hierarchical models abound \citep{Mandel2009, March2011, March2014, Rubin2015, Shariff2016, Roberts2017}, {\green and either use simulation-determined distance-corrections to correct for biases, or attempt to find analytic approximations for effects such as Malmquist bias to model the biases inside the BHM itself}.

{\green In this paper, we lay out a new hierarchical model that builds off the past work of \citet{Rubin2015}. We include additional sources of systematic uncertainty, including an analytic formulation of selection efficiency which incorporates parametric uncertainty. We also implement a different model of intrinsic dispersion to both incorporate redshift-dependent scatter and to increase numerical efficiency, allowing our model to perform rapid fits to supernovae datasets.}

Section \ref{sec:review} is dedicated to a quick review of {\green supernova cosmology analysis methods}, and Section \ref{sec:challenges} outlines some of the common challenges faced by analysis methods. In Section \ref{sec:method} we outline our methodology. Model verification on simulated datasets is given in Section \ref{sec:verification}, along with details on potential areas of improvement. We summarise our methodology in Section \ref{sec:conclusion}.

@@

Whilst supernova observations take the form of photometric time-series brightness measurements in many bands {\green and a redshift measurement of the supernova (or its assumed host)}, most analyses do not work from these measurements directly. Instead, most techniques fit an observed redshift and these photometric observations to a supernova model, with the most widely used being that of the empirical SALT2 model \citep{Guy2007, Guy2010}. This model is trained separately before fitting the supernova light curves for cosmology \citep{Guy2010, Betoule2014}. The resulting output from the model is, for each supernova, an amplitude $x_0$ (which can be converted into apparent magnitude, $m_B = -2.5\log(x_0)$), a stretch term $x_1$, and color term $c$, along with a covariance matrix describing the uncertainty on these summary statistics ($\mathcal{C_\eta}$). As all supernovae are not identical, an ensemble of supernovae form a redshift-dependent, observed population of $\hat{m}_B$, $\hat{x}_1$ and $\hat{c}$, where the hat denotes an observed variable.

This {\green ensemble of $\hat{m}_B$, $\hat{x}_1$ and $\hat{c}$} represents an observed population, which -- due to the presence of various selection effects -- may not represent the true, underlying supernova population. Accurately characterising this underlying population, its evolution over redshift, and effects from {\green host-galaxy} environment, is one of the challenges of supernova cosmology. Given some modelled underlying supernova population that lives in the redshift-dependent space $M_B$ {\green(absolute magnitude of the supernova, traditionally in the Bessell $B$ band)}, $x_1$, and $c$, the introduction of cosmology into the model is simple -- it translates the underlying population from absolute magnitude space into the observed population in apparent magnitude space. Specifically, for any given supernova our map between {\green absolute magnitude and apparent magnitude is given by the distance modulus:}

\begin{align}
\mu_{\rm obs} = m_B + \alpha x_1 - \beta c - M_B + \text{other corrections}, \label{eq:standard}
\end{align}

where $M_B$ is the mean absolute magnitude for all SN~Ia given $x_1=c=0$,  $\alpha$ is the stretch correction \citep{Phillips1993, Phillips1999}, and $\beta$ is the color correction \citep{Tripp1998} that respectively encapsulate the empirical relation that broader (longer-lasting) and bluer supernovae are brighter. {\green The distance modulus $\mu_{\rm obs}$ is a product of our observations, however a distance modulus $\mu_C$ can be precisely calculated given cosmological parameters and a redshift.} The `$\text{other corrections}$' term often includes corrections for host galaxy environment, as this has statistically significant correlations with supernova properties \citep{Kelly2010, Lampeitl2010, Sullivan2010, DAndrea2011, Gupta2011, Johansson2013, Rigault2013, Uddin2017}. In traditional $\chi^2$ analyses, Malmquist bias corrections are also included, which can take the form of a redshift-dependent function \citep{Betoule2014} or a 5D function of $c$, $x_1$, $\alpha$, $\beta$ and $z$ \citep{Kessler2017, Scolnic2017}.

@@

Traditional $\chi^2$ analyses such as that found in \citet{Riess1998, Perlmutter1999, Wood-Vasey2007, Kowalski2008, Kessler2009, Conley2011, Betoule2014}, minimise the difference in distance modulus between the observed distance modulus attained from equation \ref{eq:standard}, and the cosmologically predicted distance modulus. {\green The $\chi^2$ function minimised is}

\begin{align}
\chi^2 = (\mu_{\rm obs} - \mu_C)^\dagger C^{-1}_{\mu} (\mu_{\rm obs} - \mu_C),
\end{align}

where $C^{-1}_{\mu}$ is an uncertainty matrix that combines statistical and systematic uncertainties (see \citet{Brout18SYS} for a review of these uncertainties for the DES supernova analysis). The predicted $\mu_C$ is defined as
{\green 
\begin{align}
\mu_C &= 5 \log\left[ \frac{d_L}{10{\rm pc}} \right], \label{eq:mu0}\\
d_L& = (1+z)\frac{c}{H_0} \int_0^z \frac{dz'}{E(z')},\\
E(z) &= \sqrt{ \Omega_m (1+z')^3 + \Omega_k (1 + z')^2 + \Omega_\Lambda (1+z')^{3(1+w)}}
\end{align}
}
where $d_L$ is the luminosity distance for redshift $z$ given a specific cosmology, $H_0$ is the current value of Hubble's constant in $\kmsmpc$ and $\Omega_m$, $\Omega_k$, and $\Omega_\Lambda$ represent the energy density terms for matter, curvature and dark energy respectively.

The benefit this analysis methodology provides is speed -- for samples of hundreds of supernovae or less, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with several limitations. Firstly, formulating a $\chi^2$ likelihood results in a loss of model flexibility by assuming Gaussian uncertainty. Secondly, the method of creating a covariance matrix relies on computing partial derivatives and thus any uncertainty estimated from this method loses all information about correlation between sources of uncertainty. For example, the underlying supernova color population's mean and skewness are highly correlated, however this correlation is ignored when determining population uncertainty using numerical derivatives of population permutations. {\green Whilst correlations can be incorporated into a covariance matrix, it requires human awareness of the correlations and thus methods which can automatically capture correlated uncertainties are preferable.} Thirdly, the computational efficiency is dependent on both creating the global covariance matrix, and then inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae. A recent solution to this computational cost problem is to bin the supernovae {\green in redshift bins}, which produces a matrix of manageable size {\green and can allow fitting without matrix inversion on every step}. Whilst binning data results in some loss of information, recent works tested against simulations show that this loss of information does not create significant cosmological biases \citep{Scolnic2016, Kessler2017}.

Selection efficiency, such as the well known Malmquist bias \citep{MalmquistK.G.1922} is accounted for by correcting the determined $\mu_{\rm obs}$ from the data, or equivalently, adding a distance bias to the $\mu_C$ prediction. Specifically, Malmquist bias is the result of losing the fainter tail of the supernova population at high redshift. An example of Malmquist bias is illustrated in Figure \ref{fig:malmquist}, which simulates supernovae according to equation \eqref{eq:standard}. Simulations following survey observational strategies and geometry are used to calculate the expected bias in distance modulus, which is then {\green subtracted from} the observational data. When using traditional fitting methods such as that found in \citet{Betoule2014}, these effects are not built into the likelihood and instead are formed by correcting data. This means that the bias uncertainty is not captured fully in the $\chi^2$ distribution, and subtle correlations between cosmological or population parameters and the bias correction is lost. {\green Recent developments such as the BBC method \citep{Kessler2017} incorporate corrections dependent on $\alpha$ and $\beta$, improving their capture of uncertainty on bias corrections in the $\chi^2$ likelihood.}

@@

Sitting between the traditional models simplicity and the complexity of forward modelling lies Bayesian hierarchical models (BHM). Hierarchical models utilise multiple layers of connected parameters, with the layers linked via well defined and {\green physically} motivated conditional probabilities. For example, an observation of a parameter from a population will be conditioned on the true value of the parameter, which itself will be conditioned on the population distribution of that parameter. We can thus incorporate different population distributions, and parameter inter-dependence which cannot be found in traditional analyses where uncertainty must be encapsulated in a covariance matrix.

With the introduction of multiple layers in our model, we can add more flexibility than a traditional analysis whilst still maintaining most of the computational benefits that come from having a tractable likelihood. \citet{Mandel2009, Mandel2011a, Mandel2017} construct a hierarchical model that they apply to supernova light-curve fitting. \citet{March2011} derive a hierarchical model and simplify it by analytically marginalising over nuisance parameters to provide increased flexibility with reduced uncertainty over the traditional method, but do not incorporate bias correction. \citet{March2014} and \citet{Karpenka2015} improve upon this by incorporating redshift-dependent magnitude corrections from \citet{Perrett2010} to remove bias, and validate on 100 realisations of SNLS-like simulations. The recent BAHAMAS model \citep{Shariff2016} builds on this and reanalyses the JLA dataset \citep[using redshift dependent bias corrections from][]{Betoule2014}, whilst including extra freedom in the correction factors $\alpha$ and $\beta$, finding evidence for redshift dependence on $\beta$. \citet{Ma2016} performed a reanalysis of the JLA dataset within a Bayesian formulation, finding significant differences in $\alpha$ and $\beta$ values from the original analysis from \citet{Betoule2014}. Notably, these methods rely on data that is bias corrected or {\green the methods }ignore biases, however the UNITY framework given by \citet{Rubin2015} incorporates selection efficiency analytically in the model, and is applied to the Union 2.1 dataset \citep{Suzuki2012}. The assumption made by the UNITY analysis is that the bias {\green in real data} is {\green perfectly described} by an analytic function. {\green They validate their } model to be free of significant biases using fits to {\green thirty realisations of supernova datasets that are constructed to mimic the UNITY framework}. The well known BEAMS (Bayesian estimation applied to multiple species) methodology from \citet{Kunz2007} has been extended and applied in several works \citep{Hlozek2012}, most lately to include redshift uncertainty for photometric redshift application as zBEAMS \citep{Roberts2017} and to include simulated bias corrections in \citet{Kessler2017}. For the latter case, by inferring biases using Bayesian models, sophisticated corrections can be calculated and then applied to more traditional $\chi^2$ models.

Whilst there are a large number of hierarchical models available, none of them have undergone high-statistic tests using realistic simulations to verify each models' respective bias. {\green Additionally, testing has generally been performed on supernovae simulations with either $\Lambda$CDM cosmology or Flat $\Lambda$CDM cosmology. However, quantifying the biases on $w$CDM cosmology simulations with realistic simulations is becoming critically important as precision supernovae cosmology comes into its own, and focus shifts from determination of $\Omega_m$ to both $\Omega_m$ and $w$.}

The flexibility afforded by hierarchical models allows for investigations into different treatments of underlying supernova magnitude, color and stretch populations, host-galaxy corrections, and redshift evolution, each of which will be discussed further in the outline of our model below. Our model is designed to increase the numerical efficiency of prior works whilst incorporating the flexibility of hierarchical models. {\green We reduce our dependence on an assumed scatter model in simulations by not utilising bias-corrections in an effort} to provide a valuable cross-check on analysis methodologies which utilise scatter-model-dependant bias corrections.

@@

{\green Forefront among these challenges is our ignorance of the underlying type Ia intrinsic dispersion. Ideally, analysis of the underlying dispersion would make use of an ensemble of time-series spectroscopy to characterise the diversity of type Ia supernovae. However this data is difficult to obtain, and recent efforts to quantify the dispersion draw inference from photometric measurements. The underlying dispersion model is not a solved problem, and we therefore test against }two dispersion models in this work. The {\green first is based on the} \citet[][hereafter denoted {\gten}]{Guy2010} scatter model, which models intrinsic scatter with a 70\% contribution from coherent variation and 30\% from chromatic variation. The second, {\green based on the} {\celeven} model, is sourced from \citet{Chotard2011} and has variation with 25\% contribution from coherent scatter and 75\% from chromatic variation. {\green These two broadband scatter models are converted to spectral energy distribution models for use in simulations in \citet{Kessler13}.} As the SALT2 model does not include full treatment of intrinsic dispersion, each scatter model results in different biases in $m_B$, $x_1$, and $c$ when fitting the SALT2 model to light curve observations, and results in increased uncertainty on the summary statistics that is not encapsulated in the reported covariance $\mathcal{C_\eta}$. These two scatter models are currently {\green assumed} to span the possible range of scatter in the underlying supernova population. We have insufficient information to prefer one model over the other, and thus we have to account for both possible scatter models.

The underlying supernova population is further complicated by the presence of outliers. Non-type Ia supernovae often trigger transient follow-up in surveys and can easily be mistaken for type Ia supernovae {\green and represent outliers from the standardised SN~Ia sample.} This contamination is not just a result of non-SN~Ia being observed, but can also arise from host galaxy misidentification causing incorrect redshifts {\green being assigned to supernovae. Different optimizations to the host-galaxy algorithm can result in misidentification of the host at the 3\% to 9\% level \citep{Gupta2016}, resulting in a broad population of outliers}. For spectroscopic surveys, where both supernova type and redshift can be confirmed through the supernova spectra, this outlier population is negligible. However, for photometric surveys, which do not have the spectroscopic confirmation, it is one of the largest challenges; how to model, fit, and correct for contaminants.

Finally, one of the other persistent challenges facing supernova cosmology analyses are the high number of systematics. {\green Because of the rarity of SN~Ia events, datasets are commonly formed from the SN~Ia discoveries of multiple surveys in order to increase the number of supernovae in a dataset.} However, each different survey introduces additional sources of systematic error, from sources within each survey such as band calibration, to systematics introduced by calibration across surveys. Peculiar velocities, different host environments, and dust extinction represent additional sources of systematic uncertainty which must all be modelled and accounted for.

@@

We construct our hierarchical Bayesian model {\steve} with several goals in mind: creation of a redshift-dependent underlying supernova population, {\green treatment of an increased number of systematics}, and analytic correction of selection effects, including systematic uncertainty on those corrections. We also desire {\steve} more computationally efficient than prior works, such that cosmological results from thousands of supernovae are obtainable in the order of hours, rather than days. As this is closest to the UNITY method from \citet[][hereafter denoted \rubin]{Rubin2015}, we follow a similar model scaffold, and construct the model in the programming language Stan \citep{Carpenter2017, StanDevelopmentTeam2017}. The primary challenge of fitting hierarchical models is their large number of fit parameters, and Stan, which uses automatic differentiation and the no-U-turn Sampler (NUTS, a variant of Hamiltonian Monte Carlo), allows us to efficiently sample high dimensional parameter space.

@@

\subsubsection{Observables}
Like most of the BHM methods introduced previously, we work from the summary statistics, where each observed supernova has a brightness measurement $\hat{m}_B$ (which is analogous to apparent magnitude), stretch $\hat{x}_1$ and color $\hat{c}$, with uncertainty on those values encoded in the covariance matrix $\cov$. Additionally, each supernova has an observed redshift $\hat{z}$ and a host galaxy {\green stellar }mass associated with it, $\hat{m}$, where the mass measurement is converted into a probability of being above $10^{10}$ solar masses. We also have a probability of each supernova being a type Ia, $\hat{p}$. Our set of observables {\green input into the {\steve}} is therefore given as $\lbrace \hat{m}_B, \hat{x}_1, \hat{c}, \hat{z}, \hat{p}, \hat{m}, \cov \rbrace$, as shown in the probabilistic graphical model (PGM) in Figure \ref{fig:pgm}.
